BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets

### 2025-03-01 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
lets make some templates

--- 
title: 'BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets'
author:  |
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
    | Reed Clark Benkendorf$^1$$^,$$^2$^[Author for Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
abstract: |
  \noindent
  **Premise:** Depositing specimens to herbaria is a time consuming process. Many institutions have reduced the amount of funding for herbaria, and universities have reduced the amount of education dedicated tomaking collections and curatorial tasks. Despite this, the continual generation of herbarium specimens are essential for current and future research in evolution and ecology. In order to faciliate the continued growth of herbaria BarnebyLives was developed as tool to supplement collection notes, perform geographic and, taxonomic informatic processes, enact spell checks, produce labels, and submit digital data to increase the rate of accessioning specimens 
  **Methods and Results:** BarnebyLives uses geospatial data from the U.S. Census Bureau to provide administrative jurisdiction information, and data from other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions. It uses inhouse spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclatural innovations. Optionally the package writes driving directions to sites using Google Maps. The package outputs data in a tabular format, as well as a spatial format, for review by the user before rendering labels using LaTeX.       
  **Premise:** Depositing specimens to herbaria is a time consuming task. Many institutions have reduced the amount of funding for herbaria, and universities have reduced the amount of education dedicated to curatorial tasks and specimen deposition. Despite this, the continual generation of herbaria specimens are essential for current and future research in evolution and ecology. In order to faciliate the continued growth of herbaria BarnebyLives was developed as tool to supplement collection notes, perform geographic and, taxonomic informatic processes, enact spell checks, produce labels, and submit digital data for fast accessioning of specimens    
  **Methods and Results:** BarnebyLives uses geospatial data from the U.S. Census Bureau to provide political jurisdiction information, and data from other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions. It uses inhouse spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclatural innovations. Optionally the package writes driving directions to sites using Google Maps. The package outputs data in a tabular format, as well as a spatial format, for review by the user to accept or confirm changes, before rendering labels using LaTeX.       
  **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users the most current taxonomic opinions, generates driving directions to sites, and produces aesthetically appealing labels and shipping manifests in a matter of minutes.     
keywords: |
  herbarium, natural history museum, collections, geospatial, R, LaTeX   
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/american-journal-of-botany.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
- \usepackage[width=\textwidth]{caption}
- \usepackage{wrapfig}
- \usepackage[export]{adjustbox}
- \usepackage{rotating}
---

@@

Nearly 400 million specimens are housed in herbaria around the globe [@thiers2021herbaria].
However, The rate of accessioning new collections to herbaria diminished in the 20^th^ century as priorities in biology shifted away from describing and documenting earths biodiversity and towards understanding cellular and molecular processes [@prather2004decline; @pyke2010biological; @daru2018widespread].
This shift, among other factors, lead to a decline in the funding allocated to collections based research, the number of staff maintaining and accessioning new collections, and educating students in these practices [@funk2014erosion]. 
Historically specimens have been used to describe the taxonomic diversity of plants and document the worlds floristic diversity [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative].
However a renewed interest in herbarium data through 'big data approaches', such as museuomics, have brought herbarium collections back to the forefront of the natural sciences [@ronsted2020integrative; @marsico2020small].

Innovations in computing, specimen digitization, data sharing, DNA sequencing, and statistics have perhaps brought about greater use of herbarium specimens than ever before [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative]. 
The current uses of specimens and their ancillary data extend beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants [@ronsted2020integrative; @davis2023herbarium]. 
We anticipate that collections have yet to gain their full potential, fostered by novel approaches become more accessible, such as remote and electronic sensing, meta-barcoding, and community science  [@tosa2021rapid].
While image based or purely observational (rather than collections based) citizen science approaches (e.g. iNaturalist, BudBurst) have recently dovetailed with herbarium specimens to meet many of these research needs specimens contain rich data which are not accessible via images. 
Only specimens have the ability to: provide samples of DNA, secondary metabolites, or proteins, notes on the status and composition of the biotic and abiotic settings at time of collection, material for measuring (micro-)morphological attributes [@borges2020schrodinger], and seeds or pollen. 
These factors will ensure that specimens remain the premier botanical data source into perpetuity.

However, despite the renewed recognition of the utility of collections, efforts to grow them appear slow [@prather2004decline]. 
We conjecture this is in part because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, relative to taking photographs via professionally developed apps on smartphones [@daru2018widespread; @mishler2020spatial; @manzano2021fair]. 
While many young botanists are capable of using dichotomous keys, and other resources, to reliably identify - and able to collect satisfactory  material, we have observed that they face difficulties navigating several aspects of data acquisition and processing and preparation of labels for submission to herbaria. 
Apparent problems include the lack of dedicated time at a field seasons end to process specimens, a general lack of education on cartography and orienteering, natural history (e.g. geology, geomorphology), nomenclature, familiarity with various computer programs (e.g. Microsoft Office suite), and increasingly - foundational knowledge of plant systematics and phylogenetics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads]. 
Historically specimens have been used to describe the taxonomic diversity of plants and document the worlds floristic diversity [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative].
A renewed interest in herbarium data through 'big data approaches', such as museuomics, has brought herbarium collections back to the forefront of the natural sciences [@ronsted2020integrative; @marsico2020small].

The generation of an herbarium specimen includes many steps which are easy to take for granted [@forman1989herbarium]. 
For example while acquiring appropriate political information for a collection site appears simple, young collectors rarely have the adequate cartographic resources (printed topographic maps, or GIS software) at their disposal. 
In topographically complex areas, where administrative borders are often associated with hydrologic basins and the ridges defining them, collectors are liable to misinterpret their true geographic position and report details in error.
Even finding appropriate sites names can rarely be solved without a printed map, as many navigation related softwares, now consider many features which would serve as site names extraneous.  
Similarly, the rate at which taxonomic innovations are occurring, the volume of the literature and the reluctance of some regional curators to embrace a phylogenetic approach to plant classification, has made it difficult to find more recently applied names - even when these are unanimously accepted by group experts [@hitchcock2018flora]. 
Furthermore formatting a label correctly (e.g. author abbreviations) is a time consuming process with many opportunities to introduce several errors in formatting. 
Anecdotally, many mail merge templates offered by herbaria still require collectors to modify many variables by hand, e.g. applying italicization. 
Even if a collector navigates all of these hurdles, the time allocated to each step is quite large.

@@

Currently the label generation functionality is provided explicitly by two programs PLabel, and Symbiota, as well as commonly by the Microsoft Word tool Mail Merge (@perkins2020Plabel, @gries2014symbiota). 
The office suite costs money, and in our experience is finicky, further it's functionality ends with label creation.  
PLabel is a standalone program which has greatly enhanced functionality relative to a Mail Merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality, unfortunately it does not include data cleaning functionalities beyond verifying nations of collection; while some sources indicate it can only be used on Microsoft we expect it can be accessed on Linux and Mac using Windows emulators like Wine. 
The popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality, in an attractive GUI web portal allowing for live management of collections, and bypassing the need for a local installation, allowing it to work on all operating systems. 
Symbiota offers functionality similar to the first four of our five stages of our 'Taxonomic' module and to our knowledge a check of the 'Political Boundaries' as well (see FIG). 
However, not all herbaria use Symbiota and many have original database systems which they maintain (e.g. Harvard University Herbarium, https://kiki.huh.harvard.edu/databases/specimen_index.html, Missouri Botanical Garden https://tropicos.org/specimen/Search, and The Consortium of Pacific Northwest Herbaria https://www.pnwherbaria.org/). 
However, many collectors like to generate their own labels, especially as they are likely to be sending different sets of collections to different institutions. 
Accordingly, Symbiota's functionality should exist in an ecosystem with alternative systems. 
In scenarios where users want to keep rendering labels in either of the three existing alternatives, they can easily export data in the appropriate formats
after utilizing BLs data cleaning utilities.

BarnebyLives was named for plant taxonomist Rupert Charles Barneby (1911-2000), who published over 6,500 pages of text, described over 750 taxa, and is notable for balancing his studies at the William & Lynda Steere Herbarium at the New York Botanical Garden with annual collection trips in Western North America from 1937-1970 and sporadically until his passing in 2000 [@welsh2001rupert]. 
Select accolades of Rupert include the 1989 Asa Gray Award from the American Society of Plant Taxonomists (ASPT), the 1991 Engler Silver Medal from the International Association of Plant Taxonomists (IAPT), as well as being one of eight recipients of the International Botanical Congress's (IBC) Millennium Botany Award (1999) [@welsh2001rupert]. 
Most importantly, Rupert was remembered as being generous with his time and in assisting younger botanists with the more arcane aspects of field botany and taxonomy [@holmgren2017].

@@

![Recommended workflow](../graphics/plots/workflow.png){width=100%}

@@

## Usage
All steps of BarnebyLives except for label generation are run within the freely available Rstudio. 
Data may be read in from any common spreadsheet management system or database connection such as Excel, or free of charge alternatives such as: LibreOffice, OpenOffice, or via the cloud on Google Sheets. 
The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on the Github page (https://github.com/sagesteppe/BarnebyLives *'Input Data Column Names'*) and over 300 real-world examples are on a Google Sheets accessible from the page.
BarnebyLives is atypical of R packages in that it requires a considerable amount of data to operate (Table 1).
Virtually all the on-disk memory associated with the package are used in storing spatial data. 
The amount of spatial data will vary by the domain which the user decides to support (Figure XX). 
Functions which require the on-disk data require a path to the data as an argument. 
Manually supplying the path argument allows for users to judiciously decide a storage location suitable for their needs.

We anticipate that for the typical user BarnebyLives will requite less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S. at 90m resolution is ~16 GiB), while the processing takes relatively little RAM, hence we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
The final steps of BarnebyLives, generating the labels, requires working installations of Rmarkdown, a LaTeX installation (e.g. [pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
While these steps are run through a shell scripting language like bash, we have wrapped them in R functions which bypass the need to enter the commands directly into a terminal. 
Unfortunately we have not found Windows alternatives to pdfam and pdftek so we are unable to offer the final label generating functionality on that operating system. 
BarnebyLives was iteratively developed based on data submitted by roughly 20 seasonal field botanists. 
Essentially, additions to the code have been continually made as the developers were exposed to more idiosyncrasies of collection notes and data entry. 
Several commands in BarnebyLives require the output from previous functions, and a workflow which satisfies these requirements is presented in Figure 1.

@@

The taxonomic module has four required functions, and one optional function. 
*spell_check* will perform a spell check on the entered scientific name based on a local copy of Kew Plants of the World Database filtered to the local continents, or a user specific backbone. 
*spell_check_family* performs a spell check on the family entered for each scientific name. 
*author_check* ensures authors are entered in a valid format, e.g. the correct standard abbreviations are used. 
*associates_check* will perform a spell check on all of the associated species using the local taxonomic database.
*powo_searcher* can be used tandem to the functions *spell_check_family*, and *author-check*, but we use it in lieu of them; this function will search the current Plants of the World Online to determine relevant synonyms and alternative higher taxonomy for the focal species. 
No API key or any registration is required to use *powo_searcher*.

@@

The manual review process technically only has one function which is optional, and may be ran during the spatial process (after *coords2sf*), but the importance of manual review is important enough that it warrants explicit mention. 
*geodata_writer* will write out a spatial copy of the data set to any geospatial format supported by the sf package, but defaults to writing out 'kmls' which are readily used with [Google Earth](https://earth.google.com/web/), and can also be opened in several other free geographic information system (GIS) softwares such as [QGIS](https://www.qgis.org/). 
Notably many of the flags which BarnebyLives generates will be placed into columns (e.g. NAMES OF COLUMNS FROM FLAGS), and can be manually reviewed by the analyst, many of these issues can be resolved by simply addressing the relevant issues in the original data input spreadsheet.

The data exporting module contains three functions, which interact with LaTeX templates and require slightly more advanced R user interactivity such as setting up mapping functions using the tidyverses purrr package.
*labels_skeleton* is an R script which will require a few modification steps to tailor to each instiution, these r scripts will put data into a TEMPLATE...

@@

![The spatial extent (orange), and herbarium collection sites (burgundy) tested in this manuscript.](../graphics/plots/collections_map-trim.png){width=50%}

The package was released into beta testing using the primary authors collections from 2023. 
The testing of the package within this manuscript was performed using a subset of their collections from 2018-2022, *all* of which are un-accessioned. 
Only collections which had identifications to the level of species or lower, and transcribed collection dates and coordinates were used. 
Resulting in a data set of `r nrow(data)` records for testing, from `r n_sites` sites located across Western North America (Figure 2). 
In total `r n_spp` species (with `r n_sp_authors` distinct sets of authors), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authorships) in `r n_families` families were used for testing.

@@

```{r load in time trials}
time_trials <- readRDS('data/processed/time_trials')
tt_local <- time_trials[time_trials$Group=='Local',]

time_label_gen <- readRDS('data/processed/time_label_gen')[['elapsed']]
time_4perpage <- gsub('user.*$', '', readLines('labels/4labsperpage.txt'))[1]
time_final <- gsub('user.*$', '', readLines('labels/process2final.txt'))[1]

@@

Most of the local run time is attributable to the spatial (`r sum(tt_local[tt_local$Module=='Geospatial', 'elapsed'])`sec), and taxonomic operations (`r sum(tt_local[tt_local$Module=='Taxonomic', 'elapsed'])`sec), while formatting data for labels took `r sum(tt_local[tt_local$Module=='Style', 'elapsed'])`sec. 
The spell check operation of the scientific name accounted for nearly all of the time (`r sum(tt_local[tt_local$Function=='spell_check', 'elapsed'])`sec) spent performing local taxonomic operations.
The generation of labels consumed around `r minutemen(time_label_gen)` minutes (`r time_label_gen`sec) for the rendering, `r time_4perpage`sec to combine individual labels four per single sheet of landscape orientated paper, and `r time_final`sec to combine the `r length(list.files('labels/processed'))` sheets to a single Portable Document Format (PDF).  
The total computer run time for processing these `r nrow(data)` specimens was `r minutemen(sum(time_trials$elapsed))` minutes.

@@

`r long_flag` records were appropriately flagged for issues with auto fill increment of the longitude value, and `r lat_flag` records were also auto-flagged for increases in latitude values (% of records).

@@

![Data Sources](../graphics/tables/Screenshot-DataSources.png){width=50%}

>>>>

--- 
title: 'BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets'
author:  |
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
    | Reed Clark Benkendorf$^1$^[Author for Correspondence: rbenkendorf@chicagobotanic.org]
abstract: |
  \noindent
  **Premise:** Depositing specimens to herbaria is a time consuming process. Many institutions have reduced the amount of funding for herbaria, and universities have reduced the amount of education dedicated towards collections and curatorial tasks. However, the continual generation of herbarium specimens are essential for current and future research in evolution and ecology. In order to faciliate the continued growth of herbaria BarnebyLives was developed as a tool to supplement collection notes, perform geographic and, taxonomic informatic processes, enact spell checks and other QC steps, produce labels, and submit digital data to increase the rate of accessioning specimens.  
  **Methods and Results:** BarnebyLives uses geospatial data from the U.S. Census Bureau to provide administrative jurisdiction information, and data from other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions. It uses inhouse spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclatural innovations. Optionally the package writes driving directions to sites using Google Maps. The package outputs data in a tabular format, as well as a spatial format, for review by the user before rendering labels using LaTeX.       
  **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users the most current taxonomic opinions, generates driving directions to sites, and produces aesthetically appealing labels and shipping manifests in a matter of minutes.     
keywords: |
  herbarium, natural history museum, collections, geospatial, R, LaTeX   
output:
  word_document: default
  toc: no
  pdf_document: default
csl: "../citations/american-journal-of-botany.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
- \usepackage[width=\textwidth]{caption}
- \usepackage{wrapfig}
- \usepackage[export]{adjustbox}
- \usepackage{rotating}
---

@@

Nearly 400 million specimens are housed in herbaria around the world [@thiers2021herbaria].
However, The rate of accessioning new collections to herbaria diminished in the 20^th^ century as priorities in biology shifted away from describing and documenting earths biodiversity and towards understanding cellular and molecular processes underpinning life [@prather2004decline; @pyke2010biological; @daru2018widespread].
This shift, among other factors, lead to a decline in the funding allocated to collections based research, the number of staff maintaining and accessioning new collections, and educating students in these practices [@funk2014erosion]. 
Historically specimens have been used to describe the taxonomic diversity of plants and document the worlds floristic diversity [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative].
However, a renewed interest in herbarium collections utilizing 'big data approaches', such as museuomics, have brought herbaria back to the forefront of the natural sciences [@ronsted2020integrative; @marsico2020small].

In fact, innovations in specimen digitization, data sharing, computing, DNA sequencing, and statistics have perhaps brought about greater use of herbarium specimens than ever before [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative]. 
The current uses of specimens and their ancillary data extend well beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants [@ronsted2020integrative; @davis2023herbarium]. 
We suspect that collections have yet to realize their full potential, and as currently novel approaches, such as remote and electronic sensing and meta-barcoding, become more accessible the use of collections will increase [@tosa2021rapid].
While image based or purely observational (rather than collections based) citizen science approaches (e.g. iNaturalist, BudBurst) have recently dovetailed with herbarium specimens to meet many current research needs specimens contain rich data which are not accessible via images. 
Only specimens have the ability to: provide samples of DNA, secondary metabolites, or proteins, material for measuring (micro-)morphological attributes [@borges2020schrodinger], and seeds or pollen. 
These factors will ensure that specimens remain the premier botanical data source into perpetuity.

However, despite the renewed recognition of the utility of collections, efforts to grow them appear slow [@prather2004decline]. 
We conjecture this is in part because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, relative to taking photographs via professionally developed apps on smartphones [@daru2018widespread; @mishler2020spatial; @manzano2021fair]. 
While many young botanists are capable of using dichotomous keys, and other resources, to reliably identify - and are able to collect satisfactory  material, we have observed that they face difficulties navigating several aspects of data acquisition and processing and preparation of labels for submission to herbaria. 
Evidently, apparent problems include the lack of dedicated time at a field seasons end to process specimens, a general lack of education on cartography and orienteering, natural history (e.g. geology, geomorphology), nomenclature, familiarity with various computer programs (e.g. Microsoft Office suite), and increasingly - foundational knowledge of plant systematics and phylogenetics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].

The generation of an herbarium specimen includes many steps which are easy to take for granted [@forman1989herbarium]. 
For example while acquiring appropriate political information for a collection site appears simple, young collectors rarely have the adequate cartographic resources (printed topographic maps, or GIS software) at their disposal. 
In topographically complex areas, where administrative borders are often associated with hydrologic basins and the ridges defining them, collectors are liable to misinterpret their true geographic position and report details in error. 
Even finding appropriate sites names can rarely be solved without a printed map, as many navigation related softwares, now consider many features which would serve as site names extraneous.  
Similarly, the rate at which taxonomic innovations are occurring, the volume of the literature and the reluctance of some regional curators to embrace a phylogenetic approach to plant classification, has made it difficult to find more recently applied names - even when these names are unanimously accepted by group experts [@hitchcock2018flora]. 
Furthermore, formatting a label correctly (e.g. author abbreviations) is a time consuming process with many opportunities to introduce several errors in formatting. 
Anecdotally, many mail merge templates offered by herbaria still require collectors to modify many variables by hand, e.g. applying italicization. 
Even if a collector navigates all of these hurdles, the time allocated to each step is quite large.

@@

Currently, the label generation functionality is provided explicitly by two programs PLabel, and Symbiota, as well as commonly by the Microsoft Word tool Mail Merge (@perkins2020Plabel, @gries2014symbiota). 
The office suite costs money, and in our experience is finicky, further it's functionality ends with label creation. 
PLabel is a standalone program which has greatly enhanced functionality relative to a Mail Merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality, unfortunately it does not include data cleaning functionalities beyond verifying nations of collection.
While some sources indicate it can only be used on Microsoft we expect it can be accessed on Linux and Mac using Windows 'emulators' like Wine. 
The increasingly popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality, in an attractive GUI web portal allowing for live management of collections, and bypassing the need for a local installation, allowing it to be accessed on all operating systems. 
Symbiota offers functionality similar to the first four of our five stages of our 'Taxonomic' module and to our knowledge a check of the 'Political Boundaries' as well (see Figure 1). 
However, not all herbaria use Symbiota and many have original database systems which they maintain (e.g. Harvard University Herbarium, https://kiki.huh.harvard.edu/databases/specimen_index.html, Missouri Botanical Garden https://tropicos.org/specimen/Search, and The Consortium of Pacific Northwest Herbaria https://www.pnwherbaria.org/). 
However, many collectors like to generate their own labels, especially as they are likely to be sending different sets of collections to different institutions. 
Accordingly, Symbiota's functionality should exist in an ecosystem with alternative systems. 
In scenarios where users want to keep rendering labels in either of the three existing alternatives, they can easily export data in the appropriate formats
after utilizing BLs data cleaning utilities.

BarnebyLives was named for plant taxonomist Rupert Charles Barneby (1911-2000), who published over 6,500 pages of text, described over 750 taxa, and is notable for balancing his studies at the William & Lynda Steere Herbarium at the New York Botanical Garden with annual collection trips in Western North America from 1937-1970 and sporadically until his passing in 2000 [@welsh2001rupert]. 
Select accolades of Rupert include the 1989 Asa Gray Award from the American Society of Plant Taxonomists (ASPT), the 1991 Engler Silver Medal from the International Association of Plant Taxonomists (IAPT), as well as being one of eight recipients of the International Botanical Congress's (IBC) Millennium Botany Award (1999) [@welsh2001rupert]. 
Most germanely, Rupert was remembered as being generous with his time to assist younger botanists with the more arcane aspects of field botany and taxonomy [@holmgren2017].

@@

\includegraphics[height=\textheight, angle=270]{../graphics/plots/workflow.png}

@@

## Usage
All steps of BarnebyLives except for label generation are run within the freely available Rstudio. 
Data may be read in from any common spreadsheet management system or database connection such as Excel, or free alternatives such as: LibreOffice, OpenOffice, or via the cloud on Google Sheets. 
The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on a Github page (https://github.com/sagesteppe/BarnebyLives *'Input Data Column Names'*) and 96 real-world examples are on a Google Sheets accessible from the page.
BarnebyLives is atypical of R packages in that it requires a considerable amount of data to operate (Table 1).
Virtually all the on-disk memory associated with the package are used in storing spatial data. 
The amount of spatial data will vary by the domain which the user decides to support (Figure 3). 
Functions which require the on-disk data require a path to the data as an argument. 
Manually supplying the path argument allows for users to judiciously decide a storage location suitable for their needs.

We anticipate that for the typical user BarnebyLives will require less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S. at 90m resolution is ~16 GiB), while the processing takes relatively little RAM, hence we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
The final steps of BarnebyLives, generating the labels, requires working installations of R Markdown, a LaTeX installation (e.g. [pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
While these steps are run through a shell scripting language like bash, we have wrapped them in R functions which bypass the need to enter the commands directly into a terminal. 
Unfortunately, we have not found Windows alternatives to pdfam and pdftek so we are unable to offer the final label generating functionality on that operating system. 
BarnebyLives was iteratively developed based on data submitted by roughly 20 seasonal field botanists. 
Essentially, additions to the code have been continually made as the developers were exposed to more idiosyncrasies of collection notes and data entry.

## Usage
All steps of BarnebyLives except for label generation are run within the freely available Rstudio. 
Data may be read in from any common spreadsheet management system or database connection such as Excel, or free of charge alternatives such as: LibreOffice, OpenOffice, or via the cloud on Googlesheets. 
The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on the Github page (https://github.com/sagesteppe/BarnebyLives *'Input Data Column Names'*) and over *3*00 real-world examples are on a Google Sheets accessible from the page.
BarnebyLives is atypical of R packages in that it requires a considerable amount of data to operate (Table 1).
Virtually all the on-disk memory associated with these data are used in storing spatial data, setting up a local instance of the program - at whichever scale a user desires (see Figure XX) is fully documented in the package documentation. 
Functions which require the on-disk data require a path to the data as an argument. 
Manually supplying the path argument allows for users to judiciously decide a storage location suitable for their needs.

We anticipate BarnebyLives usage will requite less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S. at 90m resolution is ~16 GiB), and the processing takes relatively little RAM, hence we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
The final steps of BarnebyLives, generating the labels require working installations of Rmarkdown, a LaTeX installation (e.g. [pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
While these steps are run through a shell scripting language like bash, we have wrapped them in R functions which bypass the need to enter the commands directly into a terminal.
Several commands in BarnebyLives require the output from previous functions, and a workflow which satisfies these requirements is presented in Figure 1.

@@

The taxonomic module has four required functions, and one optional function. 
*spell_check* will perform a spell check on the entered scientific name based on a local copy of Kew Plants of the World database filtered to the local continents, or a user specified backbone. 
*spell_check_family* performs a spell check on the family entered for each scientific name. 
*author_check* ensures authors are entered in a valid format, e.g. the correct standard abbreviations are used. 
*associates_check* will perform a spell check on all of the associated species using the local taxonomic database.
*powo_searcher* can be used tandem to the functions *spell_check_family*, and *author-check*, but we use it in lieu of them; this function will search the current Plants of the World Online to determine relevant synonyms and alternative higher taxonomy for the focal species. 
No API key or any registration is required to use *powo_searcher*.

@@

The manual review process technically only has one function which is optional, and may be ran during the spatial process (after *coords2sf*), but the importance of manual review is important enough that it warrants explicit mention. 
*geodata_writer* will write out a spatial copy of the data set to any geospatial format supported by the sf package, but defaults to writing out 'kmls' which are readily used with [Google Earth](https://earth.google.com/web/), and can also be opened in several other free geographic information system (GIS) softwares such as [QGIS](https://www.qgis.org/). 
Notably many of the flags which BarnebyLives generates will be placed into columns with obviously flagged names, and can be manually reviewed by the analyst, many of these issues can be resolved by simply addressing the relevant issues in the original data input spreadsheet.

The data exporting module contains three functions, which interact with LaTeX templates and require slightly more advanced R user interactivity such as setting up mapping functions using the tidyverses purrr package.
*labels_skeleton* is an R 'script' which will require a few modification steps to tailor to each institution, these R scripts will put data into a user specified template, and serve as the interface to LaTeX.

@@

\begin{wrapfigure}{R}{0.35\textwidth}  
 \begin{center}
    \includegraphics[width=.35\textwidth]{../graphics/plots/collections_map-trim.png}  
  \caption{The spatial extent (orange), and herbarium collection sites (burgundy) tested in this manuscript.} 
\end{center}
\end{wrapfigure}

The package was released into beta testing using the primary authors collections from 2023. 
The testing of the package within this manuscript was performed using a subset of their collections from 2018-2022, all of which are un-accessioned. 
Only collections which had identifications to the level of species or lower, and transcribed collection dates and coordinates were used. 
Resulting in a data set of `r nrow(data)` records for testing, from `r n_sites` sites located across Western North America (Figure 2). 
In total this data set had `r n_spp` species (with `r n_sp_authors` distinct sets of authors), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authorships) in `r n_families` families were used for testing.

@@

```{r load in time trials}
time_trials <- readRDS('./data/processed/time_trials')
tt_local <- time_trials[time_trials$Group=='Local',]

time_label_gen <- readRDS('./data/processed/time_label_gen')[['elapsed']]
time_4perpage <- gsub('user.*$', '', readLines('./labels/4labsperpage.txt'))[1]
time_final <- gsub('user.*$', '', readLines('./labels/process2final.txt'))[1]

@@

Most of the local run time is attributable to the spatial (`r sum(tt_local[tt_local$Module=='Geospatial', 'elapsed'])`sec), and taxonomic operations (`r sum(tt_local[tt_local$Module=='Taxonomic', 'elapsed'])`sec), while formatting data for labels took `r sum(tt_local[tt_local$Module=='Style', 'elapsed'])`sec. 
The spell check of the scientific name accounted for nearly all of the time (`r sum(tt_local[tt_local$Function=='spell_check', 'elapsed'])`sec) spent performing local taxonomic operations.
The generation of labels consumed around `r minutemen(time_label_gen)` minutes (`r time_label_gen`sec) for the rendering, `r time_4perpage`sec to combine individual labels four per single sheet of landscape orientated paper, and `r time_final`sec to combine the `r length(list.files('labels/processed'))` sheets to a single Portable Document Format (PDF). 
The total computer run time for processing these `r nrow(data)` specimens was `r minutemen(sum(time_trials$elapsed))` minutes.

@@

*The number of author abbreviations which were not in the appropriate format were XX (% percent), in nearly all cases the presence or absence of a period were the issue.*
`r long_flag` records were appropriately flagged for issues with auto fill increment of the longitude value, and `r lat_flag` records were also auto-flagged for increases in latitude values (% of records).

@@

\begin{wrapfigure}{R}{0.45\textwidth}
  \begin{center}
    \includegraphics[width=0.40\textwidth]{../graphics/tables/Screenshot-DataSources.png}
  \caption{Sources of Data required for operations}
  \end{center}
\end{wrapfigure}
